$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-06 paper/paper.tex
on substitution

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\nitheoremautorefname}{Thm.}%
\renewcommand{\nidefinitionautorefname}{Def.}%
\renewcommand{\nilemmaautorefname}{Lem.}%
\renewcommand{\exampleautorefname}{Ex.}%
\addto\extrasUKenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\figureautorefname}{Fig.}%
}

@@

\begin{nitheorem}[Generalised substitution]
  \label{thm:subst-generalization}
  Let process $P$ be well-typed in $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  The substituted variable $i$ is capable of $m$ in $\Gamma_i$, and capable of $n$ in $\Psi_i$.
  Substitution will take these usages $m$ and $n$ away from $i$ and transfer them to the variable $j$ we are substituting for.
  In other words, there must exist some $\Gamma$, $\Psi$, $\Gamma_j$ and $\Psi_j$ such that:
  \begin{multicols}{2}
  \begin{itemize}
    \item $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$
    \item $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$
  \end{itemize}
  \end{multicols}
  Let $\Gamma$ and $\Psi$ be related such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ have a usage annotation $\lz$ at position $i$, so that all consumption from $m$ to $n$ must happen in $P$.
  Then substituting $i$ with $j$ in $P$ will be well-typed in $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  Refer to \autoref{fig:subst} for a diagramatic representation.
\end{nitheorem}

\begin{proof}[Proof (Sketch)]
  By induction on $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\constr{end}$ observe that $\Gamma_i \equiv \Psi_i$ and thus $\Gamma_j \equiv \Psi_j$.
      The details are found in \autoref{app:subst-generalization}.

    \item
      For constructor $\constr{chan}$ we proceed inductively.

    \item
      For constructors $\constr{recv}$, $\constr{send}$ and $\constr{comp}$ we must find $\Theta$ in \autoref{fig:subst} and split the diagram along its vertical axis to proceed by induction.
      The details are found in \autoref{app:subst-generalization}.
  \end{itemize}  
\end{proof}

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let $P$ be a well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof (Sketch)]
  Use framing to derive $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$.
  Use these to apply \autoref{thm:subst-generalization}.
\end{proof}

>>>>

% Do not use italics in definitions and theorems
\theoremstyle{definition}
\newtheorem{nidefinition}{Definition}
\newtheorem{nitheorem}{Theorem}
\newtheorem{nilemma}{Lemma}
\renewcommand{\nitheoremautorefname}{Thm.}%
\renewcommand{\nidefinitionautorefname}{Def.}%
\renewcommand{\nilemmaautorefname}{Lem.}%
\renewcommand{\exampleautorefname}{Ex.}%
\addto\extrasUKenglish{%
  \renewcommand{\sectionautorefname}{\S}%
  \renewcommand{\subsectionautorefname}{\S}%
  \renewcommand{\figureautorefname}{Fig.}%
  \renewcommand{\appendixautorefname}{App.}%
}

@@

\begin{nitheorem}[Generalised substitution]
  \label{thm:subst-generalization}
  Let process $P$ be well-typed in $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  The substituted variable $i$ is capable of $m$ in $\Gamma_i$, and capable of $n$ in $\Psi_i$.
  Substitution will take these usages $m$ and $n$ away from $i$ and transfer them to the variable $j$ we are substituting for.
  In other words, there must exist some $\Gamma$, $\Psi$, $\Gamma_j$ and $\Psi_j$ such that:
  \begin{multicols}{2}
  \begin{itemize}
    \item $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$
    \item $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$
    \item $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$
  \end{itemize}
  \end{multicols}
  Let $\Gamma$ and $\Psi$ be such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ at position $i$ have usage $\lz$, meaning all consumption from $m$ to $n$ must happen in $P$.
  Then substituting $i$ with $j$ in $P$ will be well-typed in $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  \autoref{fig:subst} gives a diagramatic representation.
\end{nitheorem}

\begin{proof}[Proof (Sketch)]
  By induction on $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      Constructor $\constr{end}$: observe that $\Gamma_i \equiv \Psi_i$ and thus $\Gamma_j \equiv \Psi_j$.

    \item
      Constructor $\constr{chan}$: we proceed inductively.

    \item
      Constructors $\constr{recv}$, $\constr{send}$ and $\constr{comp}$: we must find $\Theta$ in \autoref{fig:subst} and split the diagram along its vertical axis to proceed by induction.
      Full details are found in \autoref{app:subst-generalization}.
  \end{itemize}
\end{proof}

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let $P$ be a well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well-typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof (Sketch)]
  W use framing to derive $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$.
  Then, we use these to apply \autoref{thm:subst-generalization}.
\end{proof}
